Sustainable computational science:\\the ReScience initiative

### 2017-10-01 draft.tex
Adapted draft to PeerJ template (from overleaf)

\documentclass[a4paper,10pt, twocolumn]{article}

% Wider margins (compared to default)
\usepackage[margin=2.0cm]{geometry}

% For misc. computations
\usepackage{calc}
\usepackage{enumitem}

% English support (typography and hyphenation)
\usepackage[american]{babel}
\usepackage{csquotes}

% For math
\usepackage{amsmath}

% Unicode encoding
\usepackage[utf8]{inputenc}

% Color box
\usepackage{tikz,lipsum,lmodern}
\usepackage[most]{tcolorbox}

% Better default font (Libertine and Inconsolata)
\usepackage[ttscale=.875]{libertine}
\usepackage[scaled=0.96]{zi4}

%% \usepackage{mathspec}
%% \setmainfont{Minion Pro Cond}
%% \setmathsfont(Digits,Greek,Latin)[Numbers={Proportional}]{Minion Pro}
%% \setmathrm{Minion Pro}
%% \setsansfont{MyriadPro-Cond}

\usepackage{nicefrac}

% Biber
\usepackage[backend=biber,
            style=apa,
%            maxcitenames=3,
%            uniquelist=false,
%            maxbibnames=99,
%            apabackref=false,
            apamaxprtauth=99,
            natbib=true]{biblatex}
\DeclareLanguageMapping{american}{american-apa}

% Graphics
\usepackage{graphicx}

% Figure caption
\usepackage[labelsep=period]{caption}
\renewcommand{\captionfont}{\small\sffamily}
\renewcommand{\captionlabelfont}{\small\sffamily\bfseries}

% Hyperref
\usepackage{xcolor}
\definecolor{blendedblue}{rgb}{0.2, 0.2, 0.6}
\definecolor{blendedred}{rgb}{0.8, 0.2, 0.2}
\usepackage[bookmarks=true,
            breaklinks=true,
            pdfborder={0 0 0},
            citecolor=blendedblue,
            colorlinks=true,
            linkcolor=blendedblue,
            urlcolor=blendedblue,
            citecolor=blendedblue,
            linktocpage=false,
            hyperindex=true,
            linkbordercolor=white]{hyperref}
\hypersetup{colorlinks=true}

% Bibliography file
\bibliography{draft.bib}

% Headers
\usepackage{fancyhdr}
\pagestyle{fancy}
% \fancyhf{}
\rhead{\footnotesize \sf page \thepage}
\lhead{\footnotesize \sf Rougier, Hinsen et al. 2017
       \textbullet~The ReScience Initiative}
\rfoot{}
\cfoot{}
\lfoot{}

\makeatletter
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
  \textbf{\huge\@title\\}
  \vspace{5mm}
  \@author
\end{flushleft}\egroup
}
\makeatother

%% \newtcbox{\orcid}{enhanced,nobeforeafter,tcbox raise base,boxrule=0.4pt,top=0mm,bottom=0mm,
%%   right=0mm,left=4mm,arc=1pt,boxsep=2pt,before upper={\vphantom{dlg}},
%%   colframe=green!50!black,coltext=green!25!black,colback=green!10!white,
%%   overlay={\begin{tcbclipinterior}\fill[green!75!blue!50!white] (frame.south west)
%%     rectangle node[text=white,font=\sffamily\bfseries\tiny,rotate=0] {ID} ([xshift=4mm]frame.north west);\end{tcbclipinterior}}}

\newcommand{\hgap}{\hspace{1mm}}

\usepackage{tikz}
\usetikzlibrary{positioning,shapes,shadows,arrows}
\newcommand{\orcid}[1]{
    \tikz[baseline=-0.5ex]{ 
        \tikzset{lib/.style={
            rectangle split,
            rectangle split parts=2,
            rectangle split horizontal,
            rectangle split part fill={green!75!blue!50!white,green!10!white},
            rectangle split draw splits=false,
            rounded corners=1pt,
            rectangle split part align={right,right},
            draw=green!50!black,
            minimum height=8pt}
        }
        \node[lib] (var){
        \nodepart[text=green!25!black]{two} \footnotesize \href{http://orcid.org/#1}{#1}};
        \node[text=white,font=\sffamily\bfseries\footnotesize,rotate=0] at ([xshift=6pt]var.west) {ID};
    }
}

\title{Sustainable computational science:\\
       the ReScience initiative}
\author{%
  \begin{small}
    \input{authors.tex}
  \end{small}
\begin{footnotesize}
  \vspace{2mm}
  \input{affiliations.tex}
  \vspace{2mm}
  $^{\star}$ReScience co-founder
  $^{\dagger}$ReScience editor
  $^{\ddagger}$ReScience reviewer
  $^{\S}$ReScience author\\
  Corresponding author:
            \href{mailto:Nicolas.Rougier@inria.fr}{Nicolas.Rougier@inria.fr}          
  \end{footnotesize}
}

\date{}

@@

\twocolumn[
  \maketitle
  \vspace{5mm}
  \noindent
  \textbf{Abstract.}
  %
  Computer science offers a large set of tools for prototyping, writing,
  running, testing, validating, sharing and reproducing results, however
  computational science lags behind.  In the best case, authors may provide
  their source code as a compressed archive and they may feel confident their
  research is reproducible. But this is not exactly true. James Buckheit and
  David Donoho proposed more than two decades ago that an article about
  computational results is advertising, not scholarship. The actual scholarship
  is the full software environment, code, and data that produced the
  result. This implies new workflows, in particular in peer-reviews. Existing
  journals have been slow to adapt: source codes are rarely requested, hardly
  ever actually executed to check that they produce the results advertised in
  the article.
  %
  ReScience is a peer-reviewed journal that targets computational research and
  encourages the explicit replication of already published research, promoting
  new and open-source implementations in order to ensure that the original
  research can be replicated from its description. To achieve this goal, the
  whole publishing chain is radically different from other traditional
  scientific journals. ReScience resides on GitHub where each new
  implementation of a computational study is made available together with
  comments, explanations, and software tests.\\

  \textbf{Keywords:} Open Science, Computational Science,
                     Reproducibility, Replicability, Journal, Publication
  %\vspace{10mm}
  \vfill
]

\clearpage
\section*{Introduction}
There is a replication crisis in Science \citep{Baker:2016, Munaf:2017}. This
crisis has been highlighted in fields as diverse as medicine
\citep{ioannidis:2005}, psychology \citep{nosek:2015}, the political sciences
\citep{janz:2015}, and recently in the biomedical sciences
\citep{iqbal:2016}. The reasons behind such non-replicability are as diverse as
the domains in which it occurs.  In medicine, factors such as {\em study power
  and bias, the number of other studies on the same question, and importantly,
  the ratio of true to no relationships among the all relationships probed}
have been highlighted as important causes \citep{ioannidis:2005}.  In
psychology, non-replicability has been blamed on spurious p-values (p-hacking),
while in the biomedical sciences \citep{iqbal:2016}, a {\em lack of access to
  full datasets and detailed protocols for both clinical and non-clinical
  biomedical investigation} is seen as a critical factor.  The same remarks
were recently issued for chemistry \citep{coudert:2017}.  Surprisingly, the
computational sciences (in the broad sense) and computer sciences (in the
strict sense) are no exception \citep{donoho:2009,Manninen:2017} despite the
fact they rely on code and data rather than on experimental observations, which
should make them immune to the aforementioned problems.\\

When Collberg and colleagues \citep{collberg:2014, collberg:2015} decided to measure the extent of the problem
precisely, they investigated the availability of code and data as well as {\em the extent to which this code would actually build with reasonable
  effort}. The results were dramatic: of the 515 (out of 613) potentially reproducible
papers targeted by the study, the authors managed to ultimately run
only 102 (less than 20\%).
These low numbers only reflect the authors' success at running the code.
They did not check for correctness of the code (i.e., does the code actually implement what is advertised in the paper), nor the reproducibility of the results (does each run lead to the same results as in the paper).
One example of this problem can be found in \citet{topalidou:2015a}, in which the authors tried to replicate results obtained from a computational neuroscience model.
Source code was not available, neither as supplementary material to the paper nor in a public repository.
When the replicators obtained the source code after contacting the corresponding author, they found that it could not be compiled and would be difficult to reuse for other purposes.\\

@@

\begin{tcolorbox}[breakable, pad at break*=1mm,
                  colback=black!2.5, arc=0pt, outer arc=0pt, boxrule=.25pt]
\begin{footnotesize}
\textbf{Box 1.} Authors having  published in Rescience explain their motivation.\\

@@

\textbf{(\cite{viejo:2016})} Replicating previous work is a relatively routine
task every time we want to build a new model: either because we want to build
on this previous work, or because we want to compare our new model to it. We also give replication tasks to
M.Sc. students every year, as projects. In all these cases, we are confronted
with incomplete or inaccurate model descriptions, as well as with the impossibility
to obtain the original results. Contacting the original authors sometimes
solves the problem, but not so often (because of the {\em dog ate my hard
  drive} syndrome). We thus accumulate knowledge, internal to the lab, about
which model works and which doesn't, and how a given model has to be parameterized
to really work. Without any place to publish it, this knowledge is
wasted. Publishing it in ReScience, opening the discussion publicly, will be a
progress for all of us. \par
\end{footnotesize}
\end{tcolorbox}

@@

\begin{figure}
  \includegraphics[width=1.0\columnwidth]{CoScience}
  \caption{\textbf{A} The ReScience publication chain starts from an
    original research article by authors A, published in a journal, in
    conference proceedings, or as a preprint. This article constitutes
    the base material for authors B, who attempt to replicate the work
    based on its description. Success or failure to replicate is not a
    criterion for acceptance or rejection, even though failure to
    replicate requires more precaution to ensure this is not a
    misunderstanding or a bug in the new code. After review, the
    replication is published, and feedback is given to original
    authors (and editors) to inform them the work has been replicated
    (or not). \textbf{B} The CoScience proposal would require the
    replication to happen \textit{before} the actual publication. In
    case of failure, nothing will be published. In case of success,
    the publication will be endorsed by authors A and authors B with
    identified roles and will be certified as reproducible because it
    has been replicated by an independent group.}
  \label{fig:coscience}
\end{figure}

@@

% -> Computational vs Experimental: ReScience X (Etienne Roesch)

% -------------------------------------------------------------- References ---
\renewcommand*{\bibfont}{\footnotesize}
\printbibliography[title=References]

>>>>

\documentclass[fleqn,10pt,lineno]{peerj}
\usepackage{calc}
\usepackage{enumitem}
\usepackage{amsmath}

% Unicode encoding
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{hyperref}
% \usepackage{biblatex}
% \bibliography{draft}

\title{Sustainable computational science:\\the ReScience initiative}
\author[1,$\dagger$,$\S$,$\star$]{Nicolas P. Rougier}
\author[2,$\dagger$,$\star$]{Konrad Hinsen}
\author[1,$\S$]{Frédéric Alexandre}
\author[3,$\dagger$]{Thomas Arildsen}
\author[4,$\dagger$]{Lorena Barba}
\author[1,$\ddagger$]{Fabien C. Y. Benureau}
\author[5,$\dagger$]{C. Titus Brown}
\author[6,$\dagger$]{Pierre de Buyl}
\author[7,$\ddagger$]{Ozan Caglayan}
\author[8,$\ddagger$]{Andrew P. Davison}
\author[9,$\ddagger$]{Marc André Delsuc}
\author[10,$\ddagger$,$\S$]{Georgios Detorakis}
\author[11,$\S$]{Alexandra K. Diem}
\author[12,$\ddagger$]{Damien Drix}
\author[13,$\ddagger$]{Pierre Enel}
\author[14,$\dagger$,$\ddagger$,$\S$]{Benoît Girard}
\author[15,$\dagger$]{Olivia Guest}
\author[16,$\ddagger$]{Matt G. Hall}
\author[17,$\S$]{Rafael Neto Henriques}
\author[1,$\ddagger$]{Xavier Hinaut}
\author[18,$\ddagger$]{Kamil S Jaron}
\author[14,$\ddagger$,$\S$]{Mehdi Khamassi}
\author[19,$\ddagger$]{Almar Klein}
\author[20,$\ddagger$]{Tiina Manninen}
\author[21,$\ddagger$]{Pietro Marchesi}
\author[22,$\ddagger$]{Dan McGlinn}
\author[23,$\ddagger$,$\S$]{Christoph Metzner}
\author[24,$\S$]{Owen L. Petchey}
\author[25,$\ddagger$]{Hans Ekkehard Plesser}
\author[26,$\dagger$]{Timothée Poisot}
\author[27,$\dagger$]{Karthik Ram}
\author[28,$\ddagger$]{Yoav Ram}
\author[29,$\ddagger$]{Etienne Roesch}
\author[30,$\ddagger$]{Cyrille Rossant}
\author[31,$\S$]{Vahid Rostami}
\author[32,$\ddagger$,$\S$]{Aaron Shifman}
\author[33,$\S$]{Joseph Stachelek}
\author[34,$\ddagger$]{Marcel Stimberg}
\author[35,$\S$]{Frank Stollmeier}
\author[36,$\ddagger$]{Federico Vaggi}
\author[37,$\S$]{Guillaume Viejo}
\author[38,$\ddagger$,$\S$]{Julien Vitay}
\author[39,$\ddagger$]{Anya Vostinar}
\author[40,$\ddagger$]{Roman Yurchak}
\author[41,$\dagger$]{Tiziano Zito}
\affil[1]{INRIA Bordeaux Sud-Ouest Talence, France – Institut des Maladies Neurodégénératives, Université de Bordeaux, CNRS UMR 5293, Bordeaux, France – LaBRI, Université de Bordeaux, Bordeaux INP, CNRS UMR 5800, Talence, France}
\affil[2]{Centre de Biophysique Moléculaire, CNRS UPR4301, Orléans, France -- Synchrotron SOLEIL, Division Expériences, Gif sur Yvette, France}
\affil[3]{Department of Electronic Systems, Faculty of Engineering and Science, Aalborg University, Denmark}
\affil[4]{Department of Mechanical and Aerospace Engineering, the George Washington University, Washington DC, USA}
\affil[5]{Department of Population Health and Reproduction, University of California Davis, Davis, CA, USA}
\affil[6]{Instituut voor Theoretische Fysica, KU Leuven, Belgium -- PdB is a postdoctoral fellow of the Research Foundation -- Flanders (FWO)}
\affil[7]{Laboratoire d'Informatique de l'Université du Maine (LIUM), Le Mans, France}
\affil[8]{Unité de Neurosciences, Information et Complexité, CNRS FRE 3693, Gif sur Yvette, France}
\affil[9]{Institut de Génétique et de Biologie Moléculaire et Cellulaire, INSERM U964, CNRS UMR 7104, Université de Strasbourg, Illkirch, France}
\affil[10]{Department of Cognitive Sciences, University of California Irvine, USA}
\affil[11]{Computational Engineering and Design, Faculty of Engineering and the Environment, University of Southampton, UK}
\affil[12]{Department of Computer Science, Humboldt-Universität zu Berlin}
\affil[13]{Friedman Brain Institute, Icahn School of Medicine at Mount Sinai, NY, USA}
\affil[14]{Institut des Systèmes Intelligent et de Robotique, Sorbonne Universités, UPMC Univ Paris 06, CNRS, Paris, France}
\affil[15]{Experimental Psychology, University College London, UK}
\affil[16]{UCL Great Ormond St Institute of Child Health. University College London, London, UK}
\affil[17]{MRC Cognition and Brain Sciences Unit, Cambridge, UK}
\affil[18]{Department of Ecology and Evolution, University of Lausanne, Lausanne, Switzerland}
\affil[19]{Independent scholar, Enschede, The Netherlands}
\affil[20]{BioMediTech Institute and Faculty of Biomedical Sciences and Engineering, Tampere University of Technology, Tampere, Finland}
\affil[21]{Swammerdam Institute for Life Sciences, Center for Neuroscience, Faculty of Science, University of Amsterdam, Amsterdam, the Netherlands}
\affil[22]{Department of Biology, College of Charleston, Charleston, SC, USA}
\affil[23]{Centre for Computer Science and Informatics Research, University of Hertfordshire, UK}
\affil[24]{Department of Evolutionary Biology and Environmental Studies, University of Zurich, Switzerland}
\affil[25]{Faculty of Science and Technology, Norwegian University of Life Sciences, Aas, Norway -- Institute of Neuroscience and Medicine (INM-6), Jülich Research Centre, Jülich, Germany}
\affil[26]{Département de Sciences Biologiques, Université de Montréal, Montréal, Canada}
\affil[27]{Berkeley Institute for Data Science, University of California Berkeley, Berkeley, CA, USA}
\affil[28]{Department of Biology, Stanford University, Stanford, CA, USA}
\affil[29]{Centre for Integrative Neuroscience, School of Psychology, University of Reading, UK}
\affil[30]{Institute of Neurology, University College London, UK}
\affil[31]{Institute of Neuroscience \& Medicine (INM-6) and Institute for Advanced Simulation (IAS-6) -- JARA-Brain Institute I (INM-10), Jülich Research Center, Jülich, Germany}
\affil[32]{Department of Biology, University of Ottawa, Ottawa, Ontario, Canada}
\affil[33]{Department of Fisheries and Wildlife, Michigan State University, MI, USA}
\affil[34]{Sorbonne Universités, UPMC Univ Paris 06, INSERM, CNRS, Institut de la Vision, Paris, France}
\affil[35]{Network Dynamics, Max Planck Institute for Dynamics and Self-Organization, Germany}
\affil[36]{Amazon, Seattle, USA}
\affil[37]{Montreal Neurological Institute, McGill University, Montreal, Canada}
\affil[38]{Professorship for Artificial Intelligence, Department of Computer Science, Chemnitz University of Technology, Chemnitz, Germany}
\affil[39]{Department of Computer Science, Grinnell College, IA, USA}
\affil[40]{Symerio, Palaiseau, France}
\affil[41]{Neural Information Processing Group, University of Tübingen, Germany}
\affil[$^{\star}$]{ReScience co-founder}
\affil[$^{\dagger}$]{ReScience editor}
\affil[$^{\ddagger}$]{ReScience reviewer}
\affil[$^{\S}$]{ReScience author}
\corrauthor[1]{Nicolas P. Rougier}{Nicolas.Rougier@inria.fr}

\begin{abstract}
  Computer science offers a large set of tools for prototyping, writing,
  running, testing, validating, sharing and reproducing results, however
  computational science lags behind.  In the best case, authors may provide
  their source code as a compressed archive and they may feel confident their
  research is reproducible. But this is not exactly true. James Buckheit and
  David Donoho proposed more than two decades ago that an article about
  computational results is advertising, not scholarship. The actual scholarship
  is the full software environment, code, and data that produced the
  result. This implies new workflows, in particular in peer-reviews. Existing
  journals have been slow to adapt: source codes are rarely requested, hardly
  ever actually executed to check that they produce the results advertised in
  the article.  ReScience is a peer-reviewed journal that targets computational
  research and encourages the explicit replication of already published
  research, promoting new and open-source implementations in order to ensure
  that the original research can be replicated from its description. To achieve
  this goal, the whole publishing chain is radically different from other
  traditional scientific journals. ReScience resides on GitHub where each new
  implementation of a computational study is made available together with
  comments, explanations, and software tests.
\end{abstract}

@@

\flushbottom
\maketitle
\thispagestyle{empty}

\section*{Introduction}
There is a replication crisis in Science \citep{Baker:2016, Munaf:2017}. This
crisis has been highlighted in fields as diverse as medicine
\citep{ioannidis:2005}, psychology \citep{nosek:2015}, the political sciences
\citep{janz:2015}, and recently in the biomedical sciences
\citep{iqbal:2016}. The reasons behind such non-replicability are as diverse as
the domains in which it occurs.  In medicine, factors such as {\em study power
  and bias, the number of other studies on the same question, and importantly,
  the ratio of true to no relationships among the all relationships probed}
have been highlighted as important causes \citep{ioannidis:2005}.  In
psychology, non-replicability has been blamed on spurious p-values (p-hacking),
while in the biomedical sciences \citep{iqbal:2016}, a {\em lack of access to
  full datasets and detailed protocols for both clinical and non-clinical
  biomedical investigation} is seen as a critical factor.  The same remarks
were recently issued for chemistry \citep{coudert:2017}.  Surprisingly, the
computational sciences (in the broad sense) and computer sciences (in the
strict sense) are no exception \citep{donoho:2009,Manninen:2017} despite the
fact they rely on code and data rather than on experimental observations, which
should make them immune to the aforementioned problems.\\

When Collberg and colleagues \citep{collberg:2016} decided to measure the extent of the problem
precisely, they investigated the availability of code and data as well as {\em the extent to which this code would actually build with reasonable
  effort}. The results were dramatic: of the 515 (out of 613) potentially reproducible
papers targeted by the study, the authors managed to ultimately run
only 102 (less than 20\%).
These low numbers only reflect the authors' success at running the code.
They did not check for correctness of the code (i.e., does the code actually implement what is advertised in the paper), nor the reproducibility of the results (does each run lead to the same results as in the paper).
One example of this problem can be found in \citet{topalidou:2015a}, in which the authors tried to replicate results obtained from a computational neuroscience model.
Source code was not available, neither as supplementary material to the paper nor in a public repository.
When the replicators obtained the source code after contacting the corresponding author, they found that it could not be compiled and would be difficult to reuse for other purposes.\\

@@

\begin{small}
\textbf{Box 1.} Authors having  published in Rescience explain their motivation.\\

@@

\textbf{(\cite{viejo:2016})} Replicating previous work is a relatively routine
task every time we want to build a new model: either because we want to build
on this previous work, or because we want to compare our new model to it. We also give replication tasks to
M.Sc. students every year, as projects. In all these cases, we are confronted
with incomplete or inaccurate model descriptions, as well as with the impossibility
to obtain the original results. Contacting the original authors sometimes
solves the problem, but not so often (because of the {\em dog ate my hard
  drive} syndrome). We thus accumulate knowledge, internal to the lab, about
which model works and which doesn't, and how a given model has to be parameterized
to really work. Without any place to publish it, this knowledge is
wasted. Publishing it in ReScience, opening the discussion publicly, will be a
progress for all of us. \par
\end{small}

@@

\begin{figure}
  \includegraphics[width=0.75\columnwidth]{CoScience}
  \caption{\textbf{A} The ReScience publication chain starts from an
    original research article by authors A, published in a journal, in
    conference proceedings, or as a preprint. This article constitutes
    the base material for authors B, who attempt to replicate the work
    based on its description. Success or failure to replicate is not a
    criterion for acceptance or rejection, even though failure to
    replicate requires more precaution to ensure this is not a
    misunderstanding or a bug in the new code. After review, the
    replication is published, and feedback is given to original
    authors (and editors) to inform them the work has been replicated
    (or not). \textbf{B} The CoScience proposal would require the
    replication to happen \textit{before} the actual publication. In
    case of failure, nothing will be published. In case of success,
    the publication will be endorsed by authors A and authors B with
    identified roles and will be certified as reproducible because it
    has been replicated by an independent group.}
  \label{fig:coscience}
\end{figure}

@@

% -------------------------------------------------------------- References ---

\bibliography{draft}
